---
title: "Supplementary Figure 1B, SpatialPCA Results"
output: html_document
---

The following analysis uses Visium V2 data from Sample P2 in the paper [High-definition spatial transcriptomic profiling of immune cell populations in colorectal cancer](https://doi.org/10.1038/s41588-025-02193-3).

```{r, load-libraries, warning=FALSE, results="hide"}
library(rhdf5)

library(ggplot2)
library(fs)
library(glue)

library(SpatialPCA)
```

## Load the processed Data

```{r, define-helper-meta, include=FALSE}
# Helpers
read_h5ad_meta <- function(file, name='/obs'){
  cd <- rhdf5::h5read(file, name)
  
  
  cd <- purrr::map(cd, \(l) {
    if(is.list(l)) {
      cats <- l$categories
      codes <- l$codes + 1
      codes <- ifelse(codes == 0, NA, codes)
      
      out <- cats[codes]
    } else {
      out <- l
    }
    out
  })
  
  out <- data.frame(cd)
  out['index'] <- cd[['_index']]
  rownames(out) <- cd[['_index']]
  
  if (name == "/obs") {
    try({
    spc <- rhdf5::h5read(file, '/obsm/spatial')
    spc <- t(spc)
    
    out['Y'] = spc[,2]
    out['X'] = spc[,1]
  }, silent = TRUE
  )
  

  if ("in_tissue" %in% names(cd)) out[["in_tissue"]] <- as.logical(cd$in_tissue)
  if ("array_col" %in% names(cd)) {
    out[['array_col']] <- cd$array_col
    out[["array_row"]] <- cd$array_row
  }
  
  } else if(name == "/var") {
    if ("gene_ids" %in% names(cd)) out[['Symbol']] <- cd$gene_ids
  }
  
  out
}
```

```{r, define-helper-mtx, include=FALSE}
read_h5ad_matrix <- function(file) {
  
 #  mtx <- rhdf5::h5read(file,'/')
  X_ <- rhdf5::h5read(file,'/X')
  
  obs_names = rhdf5::h5read(file,'/obs/_index')
  var_names <- rhdf5::h5read(file,'/var/_index')
  
  obs_names <- as.vector(obs_names)
  var_names <- as.vector(var_names)
  
  if (is.list(X_)) {
    
    counts <- Matrix::sparseMatrix(
      i = X_$indices,
      p = X_$indptr,
      x = as.vector(X_$data),
      index1=FALSE
    )
    
  } else {
    counts <- Matrix(
      X_, 
      sparse=TRUE,
    )
  }
  
  dim_mtx <- dim(counts)
  message(
    paste('The count matrix has dimensions ', dim_mtx[1], 'x', dim_mtx[2])
  )
  
  if (length(obs_names) == dim_mtx[1]) {
    
    
    rownames(counts) <- obs_names
    colnames(counts) <- var_names
    
    print("Genes are on the columns. Transposing.... ")
    counts <- t(counts)
  } else {
    
    rownames(counts) <- var_names
    colnames(counts) <- obs_names
  }
  
  counts
}
```

```{r define-paths,include=FALSE}
pth_proc_adata <- "../data/Visium_V2_P2CRC/visium_p2crc_processed.h5ad"
```

Load the processed Visium data used in Figure 1.

```{r, load-data, include=TRUE, tidy=TRUE}

counts <- read_h5ad_matrix(pth_proc_adata)
var <- read_h5ad_meta(pth_proc_adata,"/var")
obs <- read_h5ad_meta(pth_proc_adata,"/obs")

nobs <- nrow(obs)
  if (which(dim(counts) == nobs) == 1L) {
    counts <- t(counts)
  }
```

Use same genes as in Figure 1

```{r, load-genes, message=FALSE}
# genes3k <- read_csv("../data/visium_genes3k.csv")
genes3k <- genes3k$X1
```

## Define spatialPCA object

```{r, spca-init}
spa <- new("SpatialPCA")
spa@counts <- counts[genes3k,]
spa@normalized_expr <- as.matrix(spa@counts)
spa@location <- as.matrix(obs[,c("X","Y")])
```

## Run spatial PCA

First, compute kernel

```{r, spca-kernel}
spa <- SpatialPCA_buildKernel(
  spa,
  sparseKernel = TRUE
)
```

Now, run method

```{r, spca-est-loading}
spa <- SpatialPCA_EstimateLoading(
  spa, fast=TRUE, SpatialPCnum=20)

spa <- SpatialPCA_SpatialPCs(spa, fast=TRUE)
```

Aggregate results for plotting

```{r, agg-results,include=TRUE}
spcs <- t(spa@SpatialPCs)
colnames(spcs) <- paste0("spc_r",seq(20))
obs <- cbind(obs, spcs)
```

```{r, plot-helpers, include=FALSE}
THEME_DEFAULT_LIST <- list(
  line = element_line(colour = "black", linewidth = 1,
           linetype = 1, lineend = "square"),
  rect = element_rect(fill = NA, colour = "black",
           linewidth = 1, linetype = 1),
  
  aspect.ratio = 1/1.15,
  
  panel.background = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  panel.spacing = ggplot2::unit(0, "pt"),
  
  plot.background = element_blank(),
  plot.title = element_text(
    hjust=0.5, vjust=0.5, family="Helvetica",face="bold",size=14),
  
  axis.line = element_blank(),
  axis.title = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank(),
  
  legend.position = "none"
)

theme_krpca <- function(...) {
  do.call(theme, THEME_DEFAULT_LIST) +
  theme(...)
}

```

Plot results

```{r, plot-results}
# Update for each spatial PC
pl_obs <- obs |>
  dplyr::select(X, Y, spc_r1:spc_r3) |>
  tidyr::pivot_longer(
    contains("spc"),
    names_to = "component", 
    values_to = "score"
  )

ggplot(pl_obs, aes(X,-1*Y, color=score)) +
    geom_point(size=rel(0.25), alpha=0.90) +
    facet_wrap(~ component) +
    scale_x_continuous(expand=c(0.03,0.03)) +
    scale_y_continuous(expand=c(0.03,0.03)) +
    scico::scale_color_scico(
      breaks = function(x) range(x),
      labels = c("Low", "High"),
      palette = 'vikO', direction=1,
    ) +
  theme_krpca()
```

## Session information

```{r, info}
sessionInfo()
```
